Retrofitting Effect Handlers onto OCaml

### 2020-11-20 retro-concurrency.tex
Minor edits

Next, we have the variable-sized area for the OCaml frames. In order to keep
fibers small, this area is initially 16 words in length. When the stack pointer
|rsp| becomes less than the \emph{stack threshold} (maintained in the
|Caml_state| table), the stack is said to have overflown. On stack overflow, we
copy the whole fiber to a new area with double the size. In Multicore OCaml, we
introduce stack overflow checks into the function prologue of OCaml functions.
These stack overflow operations are rare and so will be correctly predicted by
the CPU branch predictor.

@@

\vspace{-2mm}
\section{Evaluation}
\label{sec:eval}

@@

The results are presented in Table~\ref{tab:micro_noperform}. They show that
using effect handlers (concurrency monad) is 10.02$\times$ (67.09$\times$) more
expensive than the idiomatic implementation using non-tail calls. The
concurrency monad suffers due to the heap allocation of continuation frames
(which need to be garbage collected), whereas effect handlers benefit from stack
allocation of the frames. For example the number of major collections for the
\texttt{ack} benchmark is 0 for stock OCaml, 1 for \texttt{MC} and 112 for
\texttt{monad}. Our concurrency monad (and other monadic concurrency libraries
such as Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides --
exceptions, backtraces, and DWARF unwinding are no longer useful due to the lack
of a stack.

@@

\vspace{-3mm}
\section{Conclusions}
\label{sec:conc}

Our design for effect handlers in OCaml walks the tightrope of maintaining
compatibility (for profiling, debuggers and negligible overheads for existing
programs), while unlocking the full power of non-local control flow constructs.
Our evaluation shows that we have achieved our goal: we retain compatibility
with a surprisingly low performance overhead for sequential code that preserves
the spirit of ``fast exceptions'' that has always characterised OCaml
programming. We believe that the introduction of first-class fibres and effects
into OCaml -- along with a parallel runtime~\cite{Sivaramakrishnan20} -- will
open OCaml to highly task-parallel applications with minimal hit to sequential
performance.

>>>>

Next, we have the variable-sized area for the OCaml frames. In order to keep
fibers small, this area is initially 16 words in length. When the stack pointer
|rsp| becomes less than the \emph{stack threshold} (maintained in the
|Caml_state| table), the stack is said to have overflown. On stack overflow, we
copy the whole fiber to a new area with double the size. In Multicore OCaml, we
introduce stack overflow checks into the function prologue of OCaml functions.
These stack overflows are rare and so the overflow checks will be correctly predicted
by the CPU branch predictor.

@@

%\vspace{-2mm}
\section{Evaluation}
\label{sec:eval}

@@

The results are presented in Table~\ref{tab:micro_noperform}. They show that
using effect handlers (concurrency monad) is 10.02$\times$ (67.09$\times$) more
expensive than the idiomatic implementation using non-tail calls. The
concurrency monad suffers due to the heap allocation of continuation frames
(which need to be garbage collected), whereas effect handlers benefit from
stack allocation of the frames. For example, the number of major collections
for the |ack| benchmark is 0 for stock OCaml, 1 for |MC| and 112 for |monad|.
Our concurrency monad (and other monadic concurrency libraries such as
Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides -- exceptions,
backtraces, and DWARF unwinding are no longer useful due to the lack of a
stack.

@@

%\vspace{-3mm}
\section{Conclusions}
\label{sec:conc}

Our design for effect handlers in OCaml walks the tightrope of maintaining
compatibility (for profiling, debuggers and negligible overheads for existing
programs), while unlocking the full power of non-local control flow constructs.
Our evaluation shows that we have achieved our goal: we retain compatibility
with a surprisingly low performance overhead for sequential code that preserves
the spirit of ``fast exceptions'' that has always characterised OCaml
programming. We believe that the introduction of lightweight fibers to support
effect handlers into OCaml -- along with a parallel
runtime~\cite{Sivaramakrishnan20} -- will open OCaml to highly task-parallel
applications with minimal hit to sequential performance.
